Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-11-21 journal paper/content/related_work.tex
updates

The literature on learning-by-doing vs. competition \citep[\eg][]{fudenberg1983learning, dasgupta1988learning, cabral1994learning} studies firms that learn while competing against each other, so that a firm attracting more consumers reduces its per-unit production costs. However, our model focuses on firms learning product quality (as opposed to reducing production costs) and the impact that this learning has on attracting consumers.

>>>>

The literature on learning-by-doing vs. competition \citep[\eg][]{fudenberg1983learning, dasgupta1988learning, cabral1994learning} studies firms that learn while competing against each other, so that a firm attracting more consumers reduces its per-unit production costs. \gaedit{However, our model differs from these models in two important respects. The first is that in our model firms focus on learning product quality (as opposed to reducing production costs) and the second is that firms' current period actions have reputational consequences that directly impact consumer choices in future periods. These features are absent from traditional learning-by-doing models, but important components of a model that tries to capture data-based innovation.}

### 2020-11-21 journal paper/content/sec-intro.tex
updates

%We investigate these questions with
\xhdr{Our model: competition game.} We consider a stylized duopoly model in which two firms (\emph{principals}) compete for users (\emph{agents}). Principals compete on quality rather than on prices, and engage in exploration in order to learn which actions lead to high quality products. A new agent arrives and chooses a principal according to some decision rule. The principal selects an action (\eg a list of web search results), the user experiences it and reports a reward. Each principal only observes its own users. Principals commit to their strategies in advance, with the objective to maximize the market share. We investigate several model variants, %within this framework,
where we vary agents' decision rule or allow a first-mover advantage.%
\footnote{\asedit{In particular, we use different model variants (resp., Bayesian and frequentist) for theory and for simulations so as to ensure tractability. Our main findings are similar for both.}}

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a multi-armed bandit algorithm; as such, it can dynamically adjust itself to the observed rewards, but not to the competition. This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning; more on this in Section~\ref{sec:discussion}.
%Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical.
We draw on the machine learning literature to compare bandit algorithms to one another \asedit{in isolation},%
\footnote{\asedit{As opposed to ``in competition": as solutions to a stand-alone exploration problem which aims to maximize the total consumer welfare over time (see Appendix~\ref{app:bg} for self-contained background).}}
 so that we can talk about \emph{better} bandit algorithms in a principled way.
%\footnote{Such comparisons are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. More on this in Appendix~\ref{app:bg}. Generally,  ``better" algorithms are better in the long run, but could be worse initially. }
One baseline is algorithms that do not explicitly explore (\emph{greedy algorithms}); \asedit{in isolation}, they are known to be terrible for a wide variety of problem instances.

@@

\textbf{(ii)} We dilute the \HardMax agents with a small fraction of ``random agents" who choose a principal uniformly at random. We call this model \HardMaxRandom. Then better algorithms help in a big way: a sufficiently better algorithm is guaranteed to win all non-random agents after an initial learning phase. There is a caveat, however: any algorithm can be defeated by interleaving it with the greedy algorithm. This has two undesirable consequences: a better algorithm may sometimes lose in competition, and a Nash equilibrium typically does not exist.

@@

\xhdr{Interpretation: the inverted-U relationship.}
Our findings can be framed in terms of the inverted-U relationship between competition and innovation. This is a well-established concept, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better. %\citep[\eg][]{aghion2005competition,Vives-08}.
We interpret innovation as the adoption of better algorithms,%
\footnote{Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms are well-known and/or similar technologies already exist elsewhere  \citep[\eg see][]{DS-arxiv}.}
%, even if the algorithms themselves are well-known in the research literature
and control the severity of the competition by varying the agents' decision rule from \HardMax (cut-throat competition) to \HardMaxRandom to \SoftMaxRandom and all the way to the uniform selection.
%\footnote{Agents' decision rule also controls the agents' rationality. While agents' rationality and severity of competition are often modeled separately, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}
We also find another, technically different inverted-U relationship which zeroes in on the \HardMaxRandom model.
%We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.
The barriers for innovation arise entirely from the reputational consequences of exploration in competition, \asedit{without invoking monetary costs and benefits. Whereas the inverted-U relationships in prior work are due to the monetary aspects: how much can a firm invest into  R\&D and how much can it profit from the innovation.}

@@

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of observable data. Approaches such as \citet{lambrecht2015can, bajari2018impact, varian2018artificial} argue that the diminishing returns to scale and scope of data in isolation mitigate such data feedback loops,
%as non-existent
but ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly modeling the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified by the increased \emph{quality} of data gathered by better learning algorithms, not just the quantity thereof.

>>>>

%We investigate these questions with
\xhdr{Our model: competition game.} We consider a stylized duopoly model in which two firms (\emph{principals}) compete for users (\emph{agents}). Principals compete on quality rather than on prices, and engage in exploration in order to learn which actions lead to high quality products. A new agent arrives and chooses a principal according to some decision rule. The principal selects an action (\eg a list of web search results), the user experiences it and reports a reward. Each principal only observes its own users. Principals commit to their strategies in advance, with the objective to maximize \gaedit{their} market share. We investigate several model variants, %within this framework,
where we vary agents' decision rule or allow a first-mover advantage.%
\footnote{\asedit{In particular, we use different model variants (resp., Bayesian and frequentist) for theory and for simulations so as to ensure tractability. Our main findings are similar for both.}}

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a multi-armed bandit algorithm; as such, it can dynamically adjust itself to the observed rewards, but not to the competition. This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning; more on this in Section~\ref{sec:discussion}.
%Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical.
We draw on the machine learning literature to compare bandit algorithms to one another \asedit{in isolation},%
\footnote{\asedit{As opposed to ``in competition": as solutions to a stand-alone exploration problem which aims to maximize the total consumer welfare over time (see Appendix~\ref{app:bg} for self-contained background).}}
 so that we can talk about \emph{better} bandit algorithms in a principled way.
%\footnote{Such comparisons are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. More on this in Appendix~\ref{app:bg}. Generally,  ``better" algorithms are better in the long run, but could be worse initially. }
One baseline is algorithms that do not explicitly explore (\emph{greedy algorithms}); \asedit{in isolation}, they are known to \gaedit{perform poorly}\gadelete{be terrible} for a wide variety of problem instances.

@@

\textbf{(ii)} We dilute the \HardMax agents with a small fraction of ``random agents" who choose a principal uniformly at random.\gaedit{\footnote{\gaedit{We provide an interpretation of the ``random" agents as reputation-agnostic consumers that are not aware of the reputation signals or the principals' strategies.}}} We call this model \HardMaxRandom. Then better algorithms help in a big way: a sufficiently better algorithm is guaranteed to win all non-random agents after an initial learning phase. There is a caveat, however: any algorithm can be defeated by interleaving it with the greedy algorithm. This has two undesirable consequences: a better algorithm may sometimes lose in competition, and a Nash equilibrium typically does not exist.

@@

\xhdr{Interpretation: the inverted-U relationship.}
Our findings can be framed in terms of the inverted-U relationship between competition and innovation. This is a well-established concept, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better. %\citep[\eg][]{aghion2005competition,Vives-08}.
\gaedit{We recover a similar relationship even though our focus differs from the previous literature by analyzing data-based product innovation, which crucially depends on data that is generated by the consumers of the firm. We interpret the adoption of better exploration algorithms as innovation and control the severity of the competition by varying the agents' decision rule from \HardMax (cut-throat competition) to \HardMaxRandom to \SoftMaxRandom and all the way to the uniform selection.\footnote{Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms are well-known and/or similar technologies already exist elsewhere  \citep[\eg see][]{DS-arxiv}.}$^{,}$\gaedit{\footnote{\gaedit{This interpretation is what allows us to connect to the regret-minimizing bandit literature in order to discuss innovation in a principled manner by providing us with a disciplined comparison between the algorithms selected by the firm.}}}}
%\footnote{Agents' decision rule also controls the agents' rationality. While agents' rationality and severity of competition are often modeled separately, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}
We also find another, technically different inverted-U relationship which zeroes in on the \HardMaxRandom model.
%We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.
\gaedit{Unlike traditional innovation models which study lab-based innovation, our focus on data-based innovation generates the traditional inverted-U relationship purely through the reputational consequences of exploration in competition, without explicitly invoking monetary costs and benefits. This is in contrast to the inverted-U relationships established in prior work which rely on monetary aspects: how much can a firm invest into  R\&D and how much can it profit from the innovation.}

@@

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of observable data. Approaches such as \citet{lambrecht2015can, bajari2018impact, varian2018artificial} argue that the diminishing returns to scale and scope of data in isolation mitigate such data feedback loops,
%as non-existent
but ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly modeling the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified by the increased \emph{quality} of data gathered by better learning algorithms, not just the quantity thereof. \gaedit{Indeed, by exploring the first-mover advantage channel, we show how such incumbency may actually be good for innovation, but also lead to a barrier to entry precisely due to this feedback loop and how it's impacted by competition.}

### 2020-11-21 journal paper/content/sec-model.tex
updates

\asedit{\HardMaxRandom and \SoftMaxRandom can be seen as agent types that, resp., give some chance to the other principal and avoid sharp transitions in their probabilistic response. Alternatively, they can be realized as mixtures over more ``basic" agent types. Indeed, \HardMaxRandom is a mixture of \HardMax and ``random agents" (which choose a principal uniformly at random). One can obtain a \SoftMaxRandom response function using agent types that choose a principal $i$ with a largest reward estimate $\est_i$, unless $|\est_1-\est_2|$ is upper-bounded by some parameter $\theta$, in which case they choose uniformly. Then, we obtain \SoftMaxRandom as a mixture of random agents and these ``$\theta$-\HardMax'' agents, for a suitable distribution over $\theta$.}

@@

\xhdr{Competition game.}
Some of our results explicitly study the game between the two principals, termed the \emph{competition game}. We model it as a simultaneous-move game: before the first agent arrives, each principal commits to an MAB algorithm.
\asedit{Pure strategies in this game correspond to MAB algorithms.}%
\footnote{\asedit{Note that a mixed strategy is also a (randomized) MAB algorithm, \ie also a pure strategy.}}
 Principal's utility is defined as the market share, \ie the number of agents that chose this principal. Principals are risk-neutral: they optimize their expected utility.

>>>>

\asedit{\HardMaxRandom and \SoftMaxRandom can be seen as agent types that, resp., give some chance to the other principal and avoid sharp transitions in their probabilistic response. Alternatively, they can be realized as mixtures over more ``basic" agent types. Indeed, \HardMaxRandom is a mixture of \HardMax and ``random agents" (which choose a principal uniformly at random). One can obtain a \SoftMaxRandom response function using agent types that choose a principal $i$ with a largest reward estimate $\est_i$, unless $|\est_1-\est_2|$ is upper-bounded by some parameter $\theta$, in which case they choose uniformly. Then, we obtain \SoftMaxRandom as a mixture of random agents and these ``$\theta$-\HardMax'' agents, for a suitable distribution over $\theta$.} \gaedit{We provide an interpretation of the ``random agents" as reputation-agnostic consumers that are not aware of or do not seek out the reputation signals available to them in the respective choice model variants.}

@@

\xhdr{Competition game.}
Some of our results explicitly study the game between the two principals, termed the \emph{competition game}. We model it as a simultaneous-move game: before the first agent arrives, each principal commits to an MAB algorithm.
\asedit{Pure strategies in this game correspond to MAB algorithms.}%
\footnote{\asedit{Note that a mixed strategy is also a (randomized) MAB algorithm, \ie also a pure strategy.}}
 Principal's utility is defined as the market share, \ie the number of agents that chose this principal. Principals are risk-neutral \gaedit{and aim to maximize their expected utility.}\gadelete{ : they optimize their expected utility.}

### 2020-11-21 journal paper/journal_main.tex
updates

\title{\vspace{-12mm}Competing Bandits:\\
The Perils of Exploration under Competition%
\thanks{This is a merged and final version of two conference papers,
\citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19},
with a unified and streamlined presentation and expanded related work and background materials. All theoretical results are from \citet{CompetingBandits-itcs18}, and all experiments are from \citet{CompetingBandits-ec19}. Appendices~\ref{app:bg},\ref{app:examples} are completely new compared to the conference versions. We would like to thank Ian Ball, Yeon-Koo Che, Glen Weyl and Sven Rady, as well as seminar participants at Columbia and conference participants at Innovations in Theoretical Computer Science 2018 and ACM Economics and Computation 2019, for helpful comments and conversations. All errors are our own.} }
% \textit{Competing Bandits: Learning under Competition}, at Innovations in Theoretical Computer Science 2018, and \textit{The Perils of Exploration under Competition: A Computational Modeling Approach}, at ACM Economics and Computation 2019.}}

>>>>

\title{\vspace{-12mm}Competing Bandits:\\
The Perils of Exploration under Competition%
\thanks{This is a merged and final version of two conference papers,
\citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19},
with a unified and streamlined presentation and expanded related work and background materials. All theoretical results are from \citet{CompetingBandits-itcs18}, and all numerical simulations are from \citet{CompetingBandits-ec19}. Appendices~\ref{app:bg},\ref{app:examples} are completely new compared to the conference versions. We would like to thank Ian Ball, Yeon-Koo Che, Glen Weyl and Sven Rady, as well as seminar participants at Columbia and conference participants at Innovations in Theoretical Computer Science 2018, ACM Economics and Computation 2019 and the MIT Conference on Digital Experimentation 2020, for helpful comments and conversations. All errors are our own.} }
% \textit{Competing Bandits: Learning under Competition}, at Innovations in Theoretical Computer Science 2018, and \textit{The Perils of Exploration under Competition: A Computational Modeling Approach}, at ACM Economics and Computation 2019.}}
